---
title: "Some Basic Analyses with FLASH"
author: "Jean Morrison"
date: "2019-08-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The first thing I wanted to do to explore this idea was to analyze some existing data using [flashr](https://stephenslab.github.io/flashr/index.html). I used the same approach for two different applications. The first is a set of metabolic traits from different studies. The second is a set of blood cell traits from Astle et al (2016).


Mathematically the model I am fitting is this 
$$
Y = LDF^T + E
$$
where $Y$ is an $N\times M$ matrix of effect estimates ($N$ = number of variants, $M$ = number of traits), $L$ is the "loadings" matrix which is $N\timesK$ ($K$ = number of factors) and $F$ is the $M\times K$ "factors" matrix. The residuals $N\times M$ matrix $E$ is assumed to have $E_{ij} \sim N(0, s_{ij})$ where $s_{ij}$ is the estimated standard error of $Y_{ij}$.  $D$ is a  $K\times K$ diagonal matrix of weights that capture the strength of each factor. 



The procedure that I used for both data sets is this:

1. For each trait, extract all variants that have $p$-value less than $5\cdot 10^{-6}$. Take the union of these lists across traits. For each variant in the list, store the lowest $p$-value across all traits ("min $p$-value"). 
2. Prune this list of variants for LD, prioritizing based on min $p$-value. 
3. Extract three matrices. In each matrix, rows are variants and columns are traits. The three matrices store effect estimates ($\hat{\beta}$), standard errors, and $p$-values. Entries of the matrices may be missing. 
4. Run flash with a maximum of 25 factors and using both backfitting and greed algorithms. The `Y` matrix in flash is the matrix of $\hat{\beta}$ while the matrix of standard errors is the `S` matrix. The `flashr` commands used are

```{r, eval=FALSE}
data <- with(mats, flash_set_data(Y = beta_hat, S = se_hat))

f  <- flash_add_factors_from_data(data,K=kmax, var_type="zero")

f <- flash_backfit(data,f, var_type="zero")

f_greedy <- flash_add_greedy(data,Kmax=kmax, var_type="zero")
f_greedy <- flash_backfit(data,f_greedy, var_type="zero")


fits <- list(f = f, f_greedy = f_greedy)
saveRDS(fits, file=out_file)
```

So far I have only analyzed results of the backfitting algorith. These steps are implemented in a Snakemake pipeline in `/project2/xinhe/jean/gwas_factors/extract_gwas_matrix.py`.


## Metabolic Traits

First I plot a heatmap of the factors matrix. This can be interpreted as the (standardized) effect of each factor on each trait.

```{r}
library(tidyverse)
res <- readRDS("data/metabo_gwas_flash.RDS")
mats <- readRDS("data/metabo_gwas_mats.RDS")
factors <- as.data.frame(res$f$ldf$f)

traits <- str_split(mats$traits, "/") %>% map(., 8) %>% unlist(.)
traits <- str_replace(traits, "_summary_statistics.tsv.gz", "")
factors$trait <- traits
factors <- gather(factors, "factor", "value", -trait)

x <- res$f$ldf$f
rownames(x) <- traits 

heatmap(x, col=cm.colors(256))

```


## Blood Cell Traits

```{r}
library(tidyverse)
res <- readRDS("data/bc_gwas_flash.RDS")
mats <- readRDS("data/bc_gwas_mats.RDS")
factors <- as.data.frame(res$f$ldf$f)

traits <- str_split(mats$traits, "/") %>% map(., 8) %>% unlist(.)
traits <- str_replace(traits, "_summary_statistics.tsv.gz", "")
factors$trait <- traits
factors <- gather(factors, "factor", "value", -trait)

x <- res$f$ldf$f
rownames(x) <- traits 

heatmap(x, col=cm.colors(256))

```
