---
title: "Fixed Factor Correction for Row Correlation"
author: "Jean Morrison"
date: "2020-05-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


I am revisiting the idea of using fixed factors to account for sample correlation. In [previous attempts](sample_cor_ld_sims.html), that approach didn't work. Maybe the algorithm is not converging to the correct local minimum? However, the method has several advantages over the eigenvector transformaiton: 1) It is easier to deal with missing data and 2) We preserve sparsity assumptions about $F$. So I am going to try another attempt where 1) the fixed factors are scaled so that the associated loadings are all $N(0, 1)$ and second I will try a two stage fit, fitting the fixed factors and then estimating factors in the residuals.

A reminder of the model: If 

$$
Z = LF^\top + E
$$
Let $\lambda_{min}$ be the smallest eigenvalue of $\Sigma$ and $\Sigma - \lambda_{min}I = V DV^\top$

$$
E = AW^\top + E_1
$$

where elements of $E_1$ are iid N$(0, \lambda_{min})$, the elements of $A$ are iid $N(0, 1)$ and $A = V D^{1/2}$. I will try fitting on a couple examples.

```{r}
knitr::opts_chunk$set(autodep = TRUE)
library(dplyr)
library(sumstatFactors)
library(flashier)
library(ggplot2)
```

```{r, echo=FALSE}
fit_transformed <- function(dat, R, zero_thresh = 1e-15){
    R_eig <- eigen(R)
    v <- R_eig$values
    R_eig$values[abs(v) < zero_thresh] <- 0
    if(any(R_eig$values < 0))stop("R is not psd")
    d <- 1/sqrt(R_eig$values)
    d[R_eig$values==0] <- 0
    U <- R_eig$vectors %*% diag(d)
    Z_tilde <- with(dat, (beta_hat/se_beta_hat) %*% U)
    fit <- flash.init(data=Z_tilde, S = matrix(1, nrow=nrow(Z_tilde), ncol=ncol(Z_tilde)),  var.type=2) %>%
                      flash.add.greedy(Kmax = 20, init.fn = init.fn.softImpute) %>%
                      flash.backfit() %>%
                      flash.nullcheck();
    
    if(fit$n.factors > 0){
      B_hat <- fitted(fit) %*% diag(sqrt(R_eig$values)) %*% t(R_eig$vectors)
      F_hat <- R_eig$vectors %*% diag(sqrt(R_eig$values)) %*% fit$loadings.pm[[2]]
      L_hat <- fit$loadings.pm[[1]]
    }else{
      B_hat <- F_hat  <- L_hat <- NULL;
    }
  ret <- list(fit=fit, B_hat = B_hat, L_hat = L_hat, F_hat = F_hat)
  return(ret)
}

fit_naive <- function(dat){
   
    Z <- with(dat, beta_hat/se_beta_hat)
    fit <- flash.init(data=Z, S = matrix(1, nrow=nrow(Z), ncol=ncol(Z)),  var.type=2) %>%
                      flash.add.greedy(Kmax = 20, init.fn = init.fn.softImpute) %>%
                      flash.backfit() %>%
                      flash.nullcheck()
    
    if(fit$n.factors > 0){
      B_hat <- fitted(fit)
      F_hat <- fit$loadings.pm[[2]]
      L_hat <- fit$loadings.pm[[1]]
    }else{
      B_hat <- F_hat  <- L_hat <- NULL;
    }
  ret <- list(fit=fit, B_hat = B_hat, L_hat = L_hat, F_hat = F_hat)
  return(ret)
}
```

Generate some data and fit using eigenvector transformation method and navie method for comparison.

```{r}
F_mat <- readRDS("analysis_data/factors2.RDS")
F_mat[13,3] <- 0.3
set.seed(1)
R1 <- matrix(0.7, nrow=4, ncol=4)
diag(R1) <- 1
R_E <- Matrix::bdiag(R1, R1, R1, R1, R1) %>% as.matrix()
dat <- sim_sumstats_lf(F_mat = F_mat, N = 10000, J = 10000, 
                       h_2_trait = rep(0.3, 20), omega = rep(1, 20), h_2_factor = rep(0.8, 5), 
                      pi_L = rep(0.1, 5), pi_theta = 0, R_E = R_E)
#dat <- readRDS("dat.RDS")

```


```{r, cache=TRUE}
fitn <- fit_naive(dat)
fitev <- fit_transformed(dat, cov2cor(dat$Sigma))
```



```{r}
plot_factors(fitn$F_hat, 1:20)
plot_factors(fitev$F_hat, 1:20)
cor(fitn$F_hat, dat$F_mat)
cor(fitev$F_hat, dat$F_mat)
```

The eigenvector transsformation fit is clearly much better than the naive fit. This can also be seen in the RRMSE for estimated $Z$.

```{r}
B = with(dat, L_mat %*% t(F_mat) + theta)
sqrt(sum((fitn$B_hat-B)^2)/sum(B^2))
sqrt(sum((fitev$B_hat-B)^2)/sum(B^2))
```


Now fitting with fixed factors. We have found that the order of adding the fixed factors matters and whether you perform two back fits or one also matters. Also using `method = sequential` seems to work better than `extrapolate`. 


Backfitting only the fixed factors first gives the ridge regression solution since the fixed factors have a normal prior:

```{r}
R <- cov2cor(dat$Sigma)

ntrait <- ncol(dat$beta_hat)
nvar <- nrow(dat$beta_hat)
R_eig <- eigen(R)
lambda_min <- min(R_eig$values)
W <- R - lambda_min*diag(rep(1, ntrait))
W_eig <- eigen(W)
Z <- with(dat, beta_hat/se_beta_hat) 
V <- W_eig$vectors[, -ntrait]
a <- sqrt(W_eig$values[-ntrait])
Vstar <- V %*% diag(a)

#Ridge solution
Lstar <- diag(1/(a^2 + lambda_min))%*%t(Vstar)%*%t(Z) %>% t()

#flash solution
fit0 <- flash.init(Z, S =sqrt(lambda_min), var.type = NULL) %>%
          flash.init.factors(., EF = list(Lstar, Vstar), prior.family = prior.normal(scale= 1)) %>%
          flash.fix.loadings(., kset = 1:19, mode=2) %>%
          flash.backfit()
fit0$elbo

#ridge solution and flash solution are identical
cor(as.vector(Lstar %*% t(Vstar)), as.vector(fitted(fit0)))
```



If we then add additional factors, none are added. My hypothesis is that flash is stuck in a local minimum (note that here `var.type=2` rather than `NULL` so the first backfit doesn't yield exactly the ridge solution. 

```{r}
fit1 <- flash.init(Z, S =sqrt(lambda_min), var.type = 2) %>%
          flash.init.factors(., EF = list(Lstar, Vstar), prior.family = prior.normal(scale= 1)) %>%
          flash.fix.loadings(., kset = 1:19, mode=2) %>%
          flash.backfit() %>%
          flash.add.greedy() %>%
          flash.backfit()
```


If we don't backfit first and just randomly initialize the fixed factors and then add greedy factors we get only one additional factor and the elbo is slightly lower than the ridge solution. The one recovered factor corresponds the the fourth true factor. 

```{r, cache=TRUE}
set.seed(1)
L_rand <- matrix(rnorm(n=nvar*(ntrait-1)), nrow=nvar, ncol=(ntrait-1))
fit2 <- flash.init(Z, S =sqrt(lambda_min), var.type = 2) %>%
        flash.init.factors(., EF = list(L_rand, Vstar), prior.family = prior.normal(scale= 1)) %>%
        flash.fix.loadings(., kset = 1:19, mode=2) %>%
        flash.add.greedy() %>%
        flash.backfit(method="sequential", verbose.lvl=0)
fit2$elbo
F_hat <- fit2$loadings.pm[[2]][,-(1:(ntrait-1))]
cor(F_hat, dat$F_mat)
```

However, if we add the greedy factors first and then the fixed factors, we get a better solution in terms of factor recovery. The elbo is slightly worse than the fit that came up with only one factor and which we started at the ridge solution. The RRMSE is slightly worse than the RRMSE from the eigenvector transformed solution but not as bad as the RRMSE from the naive solution. 
```{r, cache=TRUE}
set.seed(2)
fit3 <-  flash.init(Z, S = sqrt(lambda_min), var.type = 2) %>%
         flash.add.greedy(Kmax = 10, init.fn = init.fn.default )

n <- fit3$n.factors

fit3 <- fit3 %>%
        flash.init.factors(., EF = list(L_rand, Vstar), prior.family = prior.normal(scale= 1)) %>%
        flash.fix.loadings(., kset = n + 1:(ntrait-1), mode=2) %>%
        flash.backfit(method = "sequential", verbose.lvl=0)
fit3$elbo
F_hat <- fit3$loadings.pm[[2]][,1:n]
plot_factors(F_hat, 1:20)
cor(F_hat, dat$F_mat)
fixed_ix <- n + (1:(ntrait-1))
B_hat <- fitted(fit3) - with(fit3, loadings.pm[[1]][, fixed_ix]%*%diag(loadings.scale[fixed_ix])%*% t(loadings.pm[[2]][, fixed_ix]))
sqrt(sum((B_hat-B)^2)/sum(B^2))
```

Trying a different seed

```{r, cache=TRUE}
set.seed(2000)
fit3 <-  flash.init(Z, S = sqrt(lambda_min), var.type = 2) %>%
         flash.add.greedy(Kmax = 10, init.fn = init.fn.default )

n <- fit3$n.factors

fit3 <- fit3 %>%
        flash.init.factors(., EF = list(L_rand, Vstar), prior.family = prior.normal(scale= 1)) %>%
        flash.fix.loadings(., kset = n + 1:(ntrait-1), mode=2) %>%
        flash.backfit(method = "sequential", verbose.lvl=0)
fit3$elbo
F_hat <- fit3$loadings.pm[[2]][,1:n]
plot_factors(F_hat, 1:20)
cor(F_hat, dat$F_mat)
fixed_ix <- n + (1:(ntrait-1))
B_hat <- fitted(fit3) - with(fit3, loadings.pm[[1]][, fixed_ix]%*%diag(loadings.scale[fixed_ix])%*% t(loadings.pm[[2]][, fixed_ix]))
sqrt(sum((B_hat-B)^2)/sum(B^2))
```

Adding only 5 (the correct number) of greedy factors:

```{r}
set.seed(2)
fit4 <-  flash.init(Z, S = sqrt(lambda_min), var.type = 2) %>%
         flash.add.greedy(Kmax = 5, init.fn = init.fn.default )

n <- fit4$n.factors

fit4 <- fit4 %>%
        flash.init.factors(., EF = list(L_rand, Vstar), prior.family = prior.normal(scale= 1)) %>%
        flash.fix.loadings(., kset = n + 1:(ntrait-1), mode=2) %>%
        flash.backfit(method = "sequential", verbose.lvl=0)
fit4$elbo
F_hat <- fit4$loadings.pm[[2]][,1:n]
plot_factors(F_hat, 1:20)
cor(F_hat, dat$F_mat)
fixed_ix <- n + (1:(ntrait-1))
B_hat <- fitted(fit4) - with(fit4, loadings.pm[[1]][, fixed_ix]%*%diag(loadings.scale[fixed_ix])%*% t(loadings.pm[[2]][, fixed_ix]))
sqrt(sum((B_hat-B)^2)/sum(B^2))
```

