---
title: "Simulating some very simple patterns"
author: "Jean Morrison"
date: "2020-08-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
This is an investigation of how FLASH fits some simple patterns. In particular, I am interested in some patterns that I have seen in real data where one factor is shared among a set of traits and another factor has opposite effects on multiple traits. This occurs in particular with inflammatory bowel disease (IBD), Chron's disease (CD), and ulcerative colitis (UC). In some fits we see one factor that affects all three and a second factor that has opposite effects on UC and and CD. An important fact is that IBD in these data is just defined as CD or UC. One guess is that in reality there is one shared factor, a set of variants that contribute to both UC and CD and then UC and CD specific variants and that the opposite sign factor is an artifact. In these simulations we explore several simple three trait patterns to see how FLASH behaves. 

```{r, include=FALSE}
knitr::opts_chunk$set(results="hide")
```


```{r, warning=FALSE, message=FALSE}
library(sumstatFactors)
library(flashier)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

## Simulation 1: One shared factor

This first pattern is a simple guess about what the IBD-CD-UC pattern might look like:

```{r}
F0 <- cbind(c(1,1,1), 
            c(1, 1, 0),
            c(0, 1, 1)) 
plot_factors(F0)
```

I set the following parameters:

+ Trait heritability: 0.5 for all traits
+ Proportion of heritability from factors: 0.7 
+ Heritability of each factor: 1 (irrelevant since we will use non-overlapping sample)
+ Proportion of non-zero elements in L_k: 0.1 for all factors (gives 1k causal variants with J = 10000)
+ Proportion of non-zero elements in theta: 0.1
+ Environmental trait correlation not mediated by factors: I (also irrelevant)
+ No LD
+ Proportion of samples overlapping: 0

### N = 10,000

We will compare FLASH fits with point normal and non-negative priors with SVD results. 
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 10000
dat <- sim_sumstats_lf(F_mat = F0, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}
xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
```

```{r}
grid.arrange(p0, p1, p2, p3, ncol=2)
```


Note that the SVD does show the opposite sign pattern that we saw in real data. FLASH finds one of the two 2-trait factors and has a small negative effect for the third trait. 


I will try modifying so the effects are weaker and there is more noise. I will change the proportion of heritabiltity that comes from factors to 0.2 (so overall the factors explain 10% of the heritability of each trait.) I will also change the proportion of non-zero elements of $\Theta$ to 0.267 so that those effects have the same size as in the previous example. 

```{r}
set.seed(5)
dat <- sim_sumstats_lf(F_mat = F0, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.2, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.27,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}
xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
```

```{r}
grid.arrange(p0, p1, p2, p3, ncol=2)
```
In this case we do a poor job of finding the shared factor but we don't get an opposite signs artifact. Next I will check with very strong signal but turning up the sample size.

### N = 100,000

```{r}
set.seed(5)
N = 100000
dat <- sim_sumstats_lf(F_mat = F0, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}

xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
```

```{r}
grid.arrange(p0, p1, p2, p3, ncol=2)
```

FLASH gets close to the right answer.

## Simulation 2: Nested

```{r}
F1 <- cbind(c(1,1,1), 
            c(1, 1, 0),
            c(0, 0, 1)) 
plot_factors(F1)
```

### N = 10,000

We will compare FLASH fits with point normal and non-negative priors with SVD results. 
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 10000
dat <- sim_sumstats_lf(F_mat = F1, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}

xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
```

```{r}
grid.arrange(p0, p1, p2, p3, ncol=2)
```

### N = 100,000
We will try increasing sample size:
```{r}
set.seed(5)
N <- 100000
dat <- sim_sumstats_lf(F_mat = F1, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}
xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p4 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
```

```{r}
grid.arrange(p0, p1, p2, p3, ncol=2)
```


## Simulation 2: Difference

```{r}
F2 <- cbind(c(1, 1, 1), c(1, -1, 0), c(0, 0, 1)) 
plot_factors(F2)
```


### N = 10,000
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 10000
dat <- sim_sumstats_lf(F_mat = F2, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}

xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
grid.arrange(p0, p1,p3, ncol=3)
```

### N = 5,000
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 5000
dat <- sim_sumstats_lf(F_mat = F2, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}

xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
grid.arrange(p0, p1, p3, ncol=3)
```


## Simulation 3: Checkers I

```{r}
F3 <- cbind(c(1, 1, 1), 
            c(1, 1, 0 ), 
            c(0, 1, 1))
plot_factors(F3)
```

### N = 10,000
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 10000
dat <- sim_sumstats_lf(F_mat = F3, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}

xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
grid.arrange(p0, p1,p2, p3, ncol=2)
```


### N = 100,000
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 100000
dat <- sim_sumstats_lf(F_mat = F3, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}
xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
grid.arrange(p0, p1,p2, p3, ncol=2)
```


## Simulation 4: Checkers II

```{r}
F4 <- cbind(c(1, 0, 1), 
            c(1, 1, 0 ), 
            c(0, 1, 1))
plot_factors(F4)
```

### N = 10,000
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 10000
dat <- sim_sumstats_lf(F_mat = F4, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}
xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
grid.arrange(p0, p1,p2, p3, ncol=2)
```

### N = 100,000
```{r}
set.seed(5)
ntrait <- 3
nfactor <- 3
nvar <- 10000
N <- 100000
dat <- sim_sumstats_lf(F_mat = F4, N = N, J = nvar, 
                       h_2_trait = rep(0.5, ntrait), omega = rep(0.7, ntrait), 
                       h_2_factor = rep(1, nfactor), pi_L = rep(0.1, nfactor), 
                       pi_theta = 0.1,
                       R_E = diag(rep(1, ntrait)), 
                       overlap_prop =0)

```

```{r}
f1 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait))
p1 <- plot_factors(f1$F_hat) + ggtitle("FLASH point-normal")

f2 <- fit_plain(dat$beta_hat, dat$se_beta_hat, rep(N, ntrait), prior_family = prior.nonnegative())
if(!is.null(f2$F_hat)){
  p2 <- plot_factors(f2$F_hat) + ggtitle("FLASH non-negative")
}else{
  p2 <- plot_factors(matrix(rep(0, 3), ncol=1)) + ggtitle("FLASH NN - no factors")
}
xx <- with(dat, beta_hat/se_beta_hat) %>% svd()
p3 <- plot_factors(xx$v)  + ggtitle("SVD")
p0 <- plot_factors(dat$F_mat, 1:5) + ggtitle("Truth")
grid.arrange(p0, p1,p2, p3, ncol=2)
```


## Bigger data

In order to see how FLASH performed when these patterns occur in a larger matrix, I constructed factor matrices for 30 traits by replicating one pattern 10 times in a block diagonal matrix. Data are simulated using the same parameters as above. For each pattern, I generated 10 different data sets. However, results were largely the same so I am only showing some example fits.

### Nested, Checckers1, Checkers2

For nested, checkers1 and checkers 2, FLASH recovers only fully shared factors. In the case of checkers2 this is interesting because in the true pattern, each factor only affects two traits. 

```{r}
res <- readRDS("analysis_data/simple_patterns/nested_1.RDS")
plot_factors(res$fit$F_hat) + ggtitle("Nested Pattern, N=10k, omega = 0.7")
```


```{r}
res <- readRDS("analysis_data/simple_patterns/checkers1_1.RDS")
plot_factors(res$fit$F_hat) + ggtitle("Checkers1 Pattern, N=10k, omega = 0.7")
```

```{r}
res <- readRDS("analysis_data/simple_patterns/checkers2_1.RDS")
plot_factors(res$fit$F_hat) + ggtitle("Checkers2 Pattern, N=10k, omega = 0.7")
```

### Difference

With the original parameter settings, FLASH is able to perfectly recover the factors:

```{r}
res <- readRDS("analysis_data/simple_patterns/difference_1.RDS")
plot_factors(res$fit$F_hat) + ggtitle("Difference Pattern, N=10k, omega = 0.7")
```
